Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2015-06-05 tex/parametrized-learning-aoas.tex
removing comments

The likelihood function is the central object that summarizes the information from an experiment needed for inference of model parameters. The likelihood function is key to Bayesian inference, and many areas of science that report the results of classical hypothesis tests or confidence intervals use the (generalized or profile) likelihood ratio as a test statistic.
%  are established tools for statistical inference~\citek{Wilks1938}. 
It is increasingly common that a simulator (or generative model) is used to describe complex processes that tie 
parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$. 
Directly evaluating the likelihood ratio in these cases is often impossible or is computationally impractical. 
Approximate Bayesian Computation (ABC) is one approach to parameter inference in this simulation-based or likelihood-free setting~\citek{Rubin1984,Tavare1997,Marin2011}. Here we consider an alternative approach that can also be used in a classical setting where a prior over the parameters is not available. In particular, we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when a generative model for the data is available for training and calibration.

@@

To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers that take advantage of a high dimensional feature vector $x$. Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\citek{Hocker:2007ht} are commonly used. Recently, there has been progress in using deep networks~\citek{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citek{HepML}, the largest Kaggle challenge in history. 
%The use of classifiers is effective, but an intermediate step not obviously connected to the target likelihood ratio test.

%Collisions are usually selected with supervised learning methods 
While classification accuracy can lead to optimal approaches for simple hypothesis tests~\citek{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters. As noted in ~\citek{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
%In what follows, it will be important to distinguish between the per-event loss used for classification, and the per-experiment loss 
The key distinction is that evaluating the loss for classification is composed of many per-event operations, while evaluating the loss for a mass measurement (e.g. the variance of an estimator for the mass parameter) is a per-experiment operation involving a data set with many events. They went on to demonstrate a computationally intensive stochastic optimization technique based on the per-experiment loss out performed the two stage selection-estimation process. 
%This paper shows that (to the extent a generalized likelihood ratio test in the full feature space is close to  optimal) one can achieve optimal results with a  specific per-event procedure by reformulating the target likelihood ratio in terms of a per-event classification followed by an specific calibration process.

@@

%In the context of machine learning or decision theory one starts with a well defined loss function.
%In order to define a Bayes risk one must also specify a prior over the parameters.
%%In practice, it can be difficult, particularly for large collaborations, to agree on such high-level topics.
%In practice, large particle physics collaborations have preferred to avoid specifying the 
%high-level loss function and prior, and instead accepted the properties of generalized likelihood ratio tests.
%Thus, we will not try to specify a loss function other than to approximate the generalized likelihood ratio itself.
%
%The form of the generalized likelihood ratio is explicitly a per-experiment object, thus naively it has the same computational challenges as other approaches that optimize a per-experiment loss. 
%This paper shows how one can reformulate generalized likelihood ratio test in terms of a per-event classification stage and a calibration process. Moreover, one can utilize a generative statistical model for the data both to train a discriminative classifier with a supervised learning algorithm and for the calibration process.
%In Section~\ref{S:Related} we discuss the scheme sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a dimensionality reduction map to aid in the estimation of the likelihood function.

%\bigskip
%
%Generalized likelihood ratio tests enjoy a number of optimal properties in the asymptotic limit and are widely used even away from that limit. These asymptotic properties, the difficulty in specifying a prior over the parameters needed to evaluate a Bayes risk, and the vast literature associated to likelihood ratio tests has led many users of such tests to not be explicit about the loss function they wish to minimize for their experiment. Thus, we will not try to specify a loss function other than to approximate the generalized likelihood ratio itself. The form of the generalized likelihood ratio is explicitly a per-experiment object, thus naively it has the same computational challenges as other approaches that optimize a per-experiment loss.

%Similarly,~\cite{Neal:2007zz} suggested using a classifier as a dimensionality reduction map to aid in the estimation of the likelihood function. We discuss the key differences with that work in Section~\ref{S:Related}.

%This paper shows how one can reformulate generalized likelihood ratio test in terms of a per-event classification stage and a calibration process. Moreover, one can utilize a generative statistical model for the data both to train a discriminative classifier with a supervised learning algorithm and for the calibration process.

%This paper describes how one can utilize generative statistical models for the data in conjunction with discriminative classifiers to construct generalized likelihood ratio tests that are equivalent to 
%the one that would have been obtained if it were possible to evaluate the probability 
%density implicitly being sampled by the generative model.
%
%This paper describes how one can utilize generative statistical models for the data in conjunction with discriminative classifiers to construct generalized likelihood ratio tests that are equivalent to 
%the one that would have been obtained if it were possible to evaluate the probability 
%density implicitly being sampled by the generative model.

%The labeled data used to train the classifiers are not obtained from the experimental apparatus. 
%Instead they are generated from a 
%detailed simulation, which includes a quantum mechanical description of the underlying particle 
%interactions, an exquisitely detailed simulation of the particle detectors with hundreds of millions of 
%sensors, and the data processing pipeline that reduces the data for a single event to tens or hundreds of high-level features used as input to the classifier. 
%It takes several minutes for this generative model to simulate a single collision or \textit{event} with Monte Carlo
%techniques. While particle physics offers an extreme example, it is common in science that generative 
%models are available while it not feasible to evaluate the probability density function that is implicitly being 
%sampled by the generative model.

@@

The first possible source of confusion we wish to avoid is that here $p(x|\theta)$  is a  \textit{generative statistical model} for the features $x$, not a generative classifier. We think of the  $p(x|\theta)$ along the lines of a traditional scientific theory or simulator, able to make predictions about $x$ and being motivated by domain-specific considerations. 
%For example, in the context of HEP $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$~\citek{Sjostrand:2006za,Agostinelli:2002hh}.  
%Moreover, we are not attempting to learn the generative model $p(x|\theta)$, we are taking it as given and trying to learn the corresponding (generalized) likelihood ratio test.
%The point here is to make the distinction of $p(x|\theta)$ in this context from a generic generative classifier like normal discriminate analysis or naive Bayes.

The second possible source of confusion is that 
we are not directly interested in calibrating the classification score in terms of a per-event posterior probability $p(y|x)$. 
Instead, we are interested in the approximation of the per-experiment likelihood function or likelihood ratio, which might be used for several purposes, including the calculation of p-values.
% defined by $P(T(D) > k |\theta)$.
%
%%the likelihood ratio $T(D)$ is aimed at tests based on the entire data set; 
%we are not interested in thresholded classification on individual events $x_e$.  
%Knowing that classifier scores are often poorly calibrated is an immediate concern.

%We wish to have well calibrated p-values defined by $P(T(D) > k |\theta)$, not well calibrated posterior probabilities $p(y|x)$.

@@

% and we need not
%interpret the balance of the training data in terms of a prior probability for the two hypotheses.
%free from the
%complications of the prior $\pi(\theta)$ that is undefined in the frequentist context.

@@

%\bigskip
%In the case of simple hypothesis testing, $\theta_0$ and $\theta_1$ are specified and there is a unique map $s(x) =  s(x_e; \theta_0, \theta_1)$. In that case, the equivalent likelihood ratio test can be performed by first transforming the data to $D_s = \{s_1, \dots, s_e\}$, constructing the likelihoods
%\begin{equation}\label{eq:NP}
%p( D_s \,|\,  \theta) = \prod_{e=1}^n \,  p( s_e \, |\,  \theta)   \; 
%\end{equation}
%for $\theta=\{\theta_0,\theta_1\}$, and constructing the likelihood ratio based on $p(D_s|\theta_0)/p(D_s|\theta_1)$.

@@

%\begin{equation}
%\frac{p(x|\theta_0)}{p(x|\theta_1)} = \frac{ p (s(x; \,\lefteqn{\overbracket[0.9pt]{\phantom{  \theta_0, \theta_1) | \,\, }}^{
%\text{identify}}} 
%\theta_0, \theta_1) | \theta_0)}
%{p (s(x; \theta_0, \,\lefteqn{\underbracket[0.9pt]{\phantom{  \theta_1) | \,\, }}_{
%\text{identify}}} 
% \theta_1) | \theta_1) } \; .
%\end{equation}

%A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citek{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
%
%In both the fixed and parametrized cases, constructing the density $\hat p(\hat s|\theta)$ requires running the generative model at $\theta$. In the context of a likelihood fit this would mean that the optimization algorithm that is trying to maximize the likelihood with respect to $\theta$ needs access to the generative model $p(x|\theta)$. This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).  In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$. 
%%In such cases, the properties of the interpolation algorithm should be part of the considerations of the over-arching optimization problem.
%One can easily imagine a number of approaches to embedding the classifier and estimating the density $\hat p(\hat s|\theta)$ and the relative merits of those approaches will depend critically on the dimensionality of $\theta$ and the computational cost of the generative model. We leave a more general strategy for this overarching optimization problem as an area of future work.

@@

%Thus far we have shown that likelihood ratio tests based on $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be reproduced via hypothesis tests based on the univariate densities $p(s|\theta)$ for the particular dimensionality reduction map $s(x|\theta_0, \theta_1)$.
% Often it is not possible to evaluate the density $p(x|\theta)$ at a given point $x$.  
%This approach will not be useful if it is not possible to approximate $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ without evaluating the density $p(x|\theta)$. 
%In order for this approach to be useful, we need to be able to approximate both based on samples $\{(x,\theta)\}$ drawn from the generative model $p(x|\theta)$.  
In order for this approach to be useful in the likelihood-free setting, where it is not possible to evaluate the density $p(x|\theta)$, we need to be able to approximate both $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.  Denote the  dimensionality reduction map $\hat{s}(x; \theta_0, \theta_1)$ and its distribution $\hat{p}(\hat{s}|\theta)$.

%In general we will be interested in two machine learning problems associated to  that approximates these distributions based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.

@@

%The first step in this direction is to confirm that a discriminative classifier obtained from common supervised learning algorithms will yield a function that is one-to-one with $p(x|\theta_0)/p(x|\theta_1)$.
%In particular, 
%is there a loss function such that the function $\hat{s}(x)$ that minimizes the expected loss leads to a function
%that is one-to-one with $p(x|\theta_0)/p(x|\theta_1)$?

@@

%Once the classifier is trained, we can use the generative model together with a univariate density estimation technique (e.g. histograms or kernel density estimation) to approximate $\hat{p}(\hat{s}|\theta)$ for specific parameter points. In order to construct an approximate likelihood function continuous in the parameters $\theta$, HEP experiments have typically used some form of interpolation of the univariate score density over the parameters $\theta$~\citek{Cranmer:2012sba}.

%The relative merits of various interpolation algorithms and the tradeoff between more samples from the generative model at a particular set of parameter points and fewer samples at more parameter points is left as a subject of future work.

%
%If we use the squared-loss function, then the expected loss is
%\begin{equation}
%\mathbb{E}[L] = \int dx p(x|\theta_0)  (  \hat{s}(x) )^2  + \int dx p(x|\theta_1)  ( 1 - \hat{s}(x) )^2 \; .
%\end{equation}
%The function $\hat{s}(x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ minimizes this expected loss.
%
%Proof (Sketch): Consider a variation about $\hat{s}(x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ given by $\hat{s}'(x) = \hat{s}(x) +h(x)$. The change in the expected loss is given by 
%\begin{equation}
%\Delta = \mathbb{E}[L_{\hat{s}'}]-\mathbb{E}[L_{\hat{s}}] = \int dx p(x|\theta_0)  (h^2(x) + 2 h(x)  \hat{s}(x) )  + p(x|\theta_1)   (h^2(x) - 2 h(x) + 2 h(x) \hat{s}(x)  \; .
%\end{equation}
%using $\hat{s}(x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ we obtain
%\begin{equation}
%\Delta =  \int dx (p(x|\theta_0)+p(x|\theta_1))  h^2(x) >0    \; .
%\end{equation}
%Thus any variation on $\hat{s}(x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ has a larger expected loss.

%The conclusion is that standard classification with a quadratic loss function and training data as described above will approximate a discriminative classifier needed to produce an equivalent likelihood ratio test.

%Once the classifier is trained, we can use the generative model and any univariate density estimation technique (e.g. histograms or kernel density estimation) to approximate $\hat{p}(\hat{s}|\theta)$.

%%% KC, commenting this out late in the game
%In the  limit of large samples from the generative model,  we can approximate the original likelihood ratio test arbitrarily well. With finite training data for $\hat{s}(x)$ and samples to approximate $\hat{p}(\hat{s}|\theta)$ it will be necessary to be more specific about the what loss function we are interested in for approximating the likelihood ratio test. This will depend in general on the ultimate goal of the test. We know that in the case of composite hypothesis tests that there is in general no uniformly most powerful test, thus it is likely that a decision theoretic approach taking into account some weighting or utility over the space $\Theta$ is necessary. This is left as a subject for future work.

@@

%A  straightforward approach to parameter inference in the likelihood-free setting is to approximate the density $\hat{p}(x|\theta)$.  In general, approximating the density $\hat{p}(x|\theta)$ is difficult for high-dimensional data, which motivates the use of the dimensionality reduction map $\hat{s}(x)$ and likelihood ratio tests based on the univariate density $\hat{p}(\hat{s}|\theta)$.  
%In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$. In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ since one does not know the true values of  $\theta_0$ and $\theta_1$; however, we can  embed the classifier as in Eq.~\ref{eq:embedding}.
% into the likelihood function to carry out the composition $\hat{p}\circ \hat{s}$. The critical observation is that  if we postpone the evaluation of the classifier to the stage of evaluating the enveloping likelihood ratio, then we can identify the parameters $\theta$ that are being compared in the likelihood ratio with the parameters used to parametrize $s(x;\theta_0, \theta_1)$.

@@

%Generalized likelihood ratio tests are used widely in HEP ~\citek{Cowan:2010js}, most notably in the discovery of the Higgs boson~\citek{Aad:2012tfa,Chatrchyan:2012ufa}.

@@

%Often machine learning classification algorithms are trained on 
In this setting, large samples of pseudo-data $\{x_i, y_i\}$ generated with some nominal values of the parameters $\nu_0$, where $y=0$ corresponds to the background density $p_b(x|\nu_0)$  and $y=1$ corresponds to signal density $p_s(x|\nu_0)$. Importantly, the $y=1$ label corresponds to the signal only, and not to the alternate signal-plus-background hypothesis. The resulting classifier approximates the regression function $p_s(x|\nu_0)/(p_s(x|\nu_0)+p_b(x|\nu_0))$, which is one to one with the likelihood ratio of the null to the alternate $p(x|\mu=0,\nu_0)/p(x|\mu,\nu_0)$ for all $\mu$. Associating the $y=1$ label to the signal component has advantages because it helps the classifier focus its capacity on the relevant regions in the feature space, particularly when the signal is a very small perturbation to the background (ie. $\mu \ll 1$). 
%At the LHC one can have $\mu \sim 10^{-6}$. 
%Note that the classifier is not optimal for $\nu\ne\nu_0$.

@@

These steps feed into a subsequent statistical test based on the observed data 
${D=(x_1, \dots, x_n)}$. For each event, the classifier is evaluated and one performs inference on a parameter $\mu$ related to the presence of the signal contribution. In particular, one forms the statistical model\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known, which is referred to as an extended likelihood or marked Poisson model.} 
\begin{equation}\label{eq:typicalML}
p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu \hat{p}_s( \hat s(x_e) \, |\,  \nu)  + (1-\mu)\, \hat{p}_b( \hat s(x_e) \,|\, \nu) \,\right] \; .
\end{equation}
%where $\mu=0$ is the null (background-only) hypothesis and $\mu>0$ is the alternate (signal-plus-background) hypothesis.\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known, which is referred to as an extended likelihood.} 
%Typically, we are interested in inference on $\mu$ and $\nu$ are nuisance parameters.

@@

There is an interesting distinction between this approach and the standard use in which the classifier is trained for a fixed $\nu_0$. In the standard use one trains a classifier for signal vs. background, which is equivalent (in an ideal setting) to training a classifier for  null (background-only) vs. alternate (signal-plus-backgound) since the resulting regression functions are one-to-one with each other.
% as 
%\begin{equation}
% \frac{p(x| 0, \nu_0)}{p(x|\hat \mu, \nu_0)} =  \frac{p_b(x| \nu_0)}{ \mu p_s( x_e \, |\,  \nu_0)  + (1-\mu)\, p_b( x_e \,|\, \nu_0)} = \left[ c_1 + c_2   \frac{p_s(x| \nu_0)}{ p_b( x_e \,|\, \nu_0)} \right ]^{-1} \; ,
%\end{equation}
%and $c_1$ and $c_2$ are constants. Specifically, the two  likelihood ratios are in one-to-one correspondence, so an ideal algorithm would lead to equivalent tests. 
In contrast, in the case of the generalized likelihood ratio test 
\begin{equation}\label{eq:hep_improved}
 \frac{p(x| 0, \hat{\hat{ \nu}})}{p(x|\hat \mu, \hat\nu)} =  \frac{p_b(x| \hat{\hat{ \nu}})}{ \hat \mu p_s( x_e \, |\,  \hat\nu)  + (1- \hat \mu )\, p_b( x_e \,|\, \hat \nu)} \; ,
\end{equation}
the background components don't cancel and there is an additional term $p_b(x| \hat{\hat{ \nu}})/p_b(x| {\hat{ \nu}})$.
In practice, with classifiers of finite capacity, there will be some tradeoff between taking into account this additional term and the more challenging learning problem when $\mu$ is very small.

@@

In this section we generalize the capacity focusing technique of training classifiers to discriminate between components of a mixture model. First, we generalize Eq.~\ref{eq:hepGen} to a mixture model of several components
\begin{equation}
p(x|\theta)=\sum_c w_c(\theta) p_c(x| \theta) \;.
\end{equation}
%In the case of particle physics, the distributions $p(x|\theta)$ is not a Gaussian Mixture Model, but mixture of  complicated distributions associated to relatively few types of particle interactions. Moreover, when searching for a new particle, the null hypothesis would correspond to some of the coefficients $w_c=0$ while the alternate ``signal-plus-background'' hypothesis would have $0<w_{c\in \rm signal} \ll w_{c\in \rm background}$. In some cases $w_{c\in \rm signal} / w_{c\in \rm background} <10^{-6}$, which means the alternate hypothesis is a small perturbation to the null hypothesis. This can be a challenge for typical classifiers because they should devote their capacity to the region where $p_{c\in \rm signal}(x) / p_{c\in \rm background}(x)$ is relatively large. Lastly, even when the distributions $p_c(x|\theta)$ are well known, it is often the case that the coefficients are uncertain or treated as completely unknown. These all present challenges to machine learning algorithms that aim to learn $s(x; \theta_0, \theta_1)$.
%
%\begin{equation}
%\frac{p(x|\theta_0)}{p(x|\theta_1)}= \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_c w_c(\theta_1) p_c(x| \theta_1)} 
%\end{equation}
%
%\begin{equation}
%\frac{p(x|\theta_0)}{p(x|\theta_1)}= \sum_c \frac{ w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} 
%\end{equation}
%
It is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems. 
\begin{eqnarray}\label{eq:decompose}
\frac{p(x|\theta_0)}{p(x|\theta_1)} & =& \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} 
\\
%&=& \sum_c \frac{ w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} \\
%&=& \sum_c \left[ \frac{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)}{ w_c(\theta_0) p_c(x| \theta_0)}  \right]^{-1} \\
&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(x| \theta_1)}{  p_c(x| \theta_0)}  \right]^{-1} \\ \label{eq:decomposedResult}
&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ p_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} 
\end{eqnarray}
The second line is a trivial, but useful decomposition into pair-wise classification between $p_{c'}(x|\theta_1)$ and $p_c(x|\theta_0)$.  The third line uses Theorem~1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$. In the situation where the only free parameters of the  model are the mixture coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before evaluating the  likelihood ratio). Equation~\ref{eq:decomposedResult} allows one to take advantage of both the parametrized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.

%
%The idea here is to combine the calibration of the distributions of the classifier output and a more optimal family of classifiers $\hat s(x; \nu_0, \nu_1)$. Creating the family of classifiers is straight forward, one simply augments the training data with $x$ examples drawn from several values of $\nu$ and then includes the corresponding value of $\nu$ as an input to the classifier. Thus $\{x_e,c_e\} \to \{x_e,\nu_{0,e},\nu_{1,e}, c_e\}$ leading to a parametrized learner $\hat s(x)\to \hat s(x;\nu_0, \nu_1)$. This leads to a complication: one does not know the value of $\nu$ to use when evaluating the parametrized classifier $\hat s(x;\nu_0, \nu_1)$, so one cannot simply pre-compute $\hat s_e$ before performing the likelihood fit.  However, when performing a likelihood ratio test, one can calculate the maximum likelihood estimate and conditional maximum likelihood estimate according to Eq.~\ref{eq:mle} and then form the equivalent likelihood ratio test according to Eq.~\ref{eq:equivLRtest}.

%Equivalently, once one has determined $\hat{\theta}$ and $\hat{\hat{\theta}}$, then one can specify the parameters of the mapping $s(x;\hat{\hat{\theta}}, \hat{\theta})$ and then following Eq.~\ref{eq:NP} form the likelihood
%\begin{equation}\label{eq:parametrizedML}
%p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( s(x_e;\nu_0, \nu_1) \, |\,  \nu)  + (1-\mu)\, p_b( s(x_e;\nu_0, \nu_1) \,|\, \nu) \,\right] \; ,
%\end{equation}
%which is a natural generalization of Eq.~\ref{eq:typicalML}.

%The construction in Eq.~\ref{eq:parametrizedML} presents a computational challenge as for each value of $\nu$ one has a new mapping $s: x\to\mathbb{R}$. The most naive realization of this would be to generate synthetic data $\{x_e\}$ according to $p_c(x|\nu)$, evaluate $s(x_e ; \nu_0, \nu_1)$, and then use a histogram, kernel density, or other non-parametric density estimation procedure to estimate $p_c( s | \nu)$. This approach is not realistic in situations where $p_c(x|\nu)$ is realized with expensive computer simulations; thus, some interpolation strategy or emulator over a fixed set of parameter points $\{\nu_i\}$ would typically be used.

@@

Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\citek{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\citek{Allanach:2000kt}.  Both of these examples involve data sets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters. 
%The state of the art approaches for supersymmetric particle masses typically involve high-level features engineered based on physics intuition
%Many approaches have been tried for measuring supersymmetric particle masses, often based on event level features engineered based on physics intuition and varying 
The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citek{Gao:2010qx,Bolognesi:2012mm}.

@@

%Refs to include:
%\citek{ClaytonScott,AndrewY.Ng,SamT.Roweis,EricP.Xing,SujaySanghavi,BiancaZadrozny,Zadrozny2001,}

>>>>

The likelihood function is the central object that summarizes the information from an experiment needed for inference of model parameters. The likelihood function is key to Bayesian inference, and many areas of science that report the results of classical hypothesis tests or confidence intervals use the (generalized or profile) likelihood ratio as a test statistic.
It is increasingly common that a simulator (or generative model) is used to describe complex processes that tie 
parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$. 
Directly evaluating the likelihood ratio in these cases is often impossible or is computationally impractical. 
Approximate Bayesian Computation (ABC) is one approach to parameter inference in this simulation-based or likelihood-free setting~\citek{Rubin1984,Tavare1997,Marin2011}. Here we consider an alternative approach that can also be used in a classical setting where a prior over the parameters is not available. In particular, we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when a generative model for the data is available for training and calibration.

@@

To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers that take advantage of a high dimensional feature vector $x$. Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\citek{Hocker:2007ht} are commonly used. Recently, there has been progress in using deep networks~\citek{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citek{HepML}, the largest Kaggle challenge in history.

While classification accuracy can lead to optimal approaches for simple hypothesis tests~\citek{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters. As noted in ~\citek{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
The key distinction is that evaluating the loss for classification is composed of many per-event operations, while evaluating the loss for a mass measurement (e.g. the variance of an estimator for the mass parameter) is a per-experiment operation involving a data set with many events. They went on to demonstrate a computationally intensive stochastic optimization technique based on the per-experiment loss out performed the two stage selection-estimation process.

@@

@@

The first possible source of confusion we wish to avoid is that here $p(x|\theta)$  is a  \textit{generative statistical model} for the features $x$, not a generative classifier. We think of the  $p(x|\theta)$ along the lines of a traditional scientific theory or simulator, able to make predictions about $x$ and being motivated by domain-specific considerations.

The second possible source of confusion is that 
we are not directly interested in calibrating the classification score in terms of a per-event posterior probability $p(y|x)$. 
Instead, we are interested in the approximation of the per-experiment likelihood function or likelihood ratio, which might be used for several purposes, including the calculation of p-values.

@@

@@

@@

@@

In order for this approach to be useful in the likelihood-free setting, where it is not possible to evaluate the density $p(x|\theta)$, we need to be able to approximate both $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.  Denote the  dimensionality reduction map $\hat{s}(x; \theta_0, \theta_1)$ and its distribution $\hat{p}(\hat{s}|\theta)$.

@@

@@

@@

@@

@@

In this setting, large samples of pseudo-data $\{x_i, y_i\}$ generated with some nominal values of the parameters $\nu_0$, where $y=0$ corresponds to the background density $p_b(x|\nu_0)$  and $y=1$ corresponds to signal density $p_s(x|\nu_0)$. Importantly, the $y=1$ label corresponds to the signal only, and not to the alternate signal-plus-background hypothesis. The resulting classifier approximates the regression function $p_s(x|\nu_0)/(p_s(x|\nu_0)+p_b(x|\nu_0))$, which is one to one with the likelihood ratio of the null to the alternate $p(x|\mu=0,\nu_0)/p(x|\mu,\nu_0)$ for all $\mu$. Associating the $y=1$ label to the signal component has advantages because it helps the classifier focus its capacity on the relevant regions in the feature space, particularly when the signal is a very small perturbation to the background (i.e. $\mu \ll 1$).

@@

These steps feed into a subsequent statistical test based on the observed data 
${D=(x_1, \dots, x_n)}$. For each event, the classifier is evaluated and one performs inference on a parameter $\mu$ related to the presence of the signal contribution. In particular, one forms the statistical model\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known, which is referred to as an extended likelihood or marked Poisson model.} 
\begin{equation}\label{eq:typicalML}
p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu \hat{p}_s( \hat s(x_e) \, |\,  \nu)  + (1-\mu)\, \hat{p}_b( \hat s(x_e) \,|\, \nu) \,\right] \; .
\end{equation}

@@

There is an interesting distinction between this approach and the standard use in which the classifier is trained for a fixed $\nu_0$. In the standard use one trains a classifier for signal vs. background, which is equivalent (in an ideal setting) to training a classifier for  null (background-only) vs. alternate (signal-plus-backgound) since the resulting regression functions are one-to-one with each other.
In contrast, in the case of the generalized likelihood ratio test 
\begin{equation}\label{eq:hep_improved}
 \frac{p(x| 0, \hat{\hat{ \nu}})}{p(x|\hat \mu, \hat\nu)} =  \frac{p_b(x| \hat{\hat{ \nu}})}{ \hat \mu p_s( x_e \, |\,  \hat\nu)  + (1- \hat \mu )\, p_b( x_e \,|\, \hat \nu)} \; ,
\end{equation}
the background components don't cancel and there is an additional term $p_b(x| \hat{\hat{ \nu}})/p_b(x| {\hat{ \nu}})$.
In practice, with classifiers of finite capacity, there will be some tradeoff between taking into account this additional term and the more challenging learning problem when $\mu$ is very small.

@@

In this section we generalize the capacity focusing technique of training classifiers to discriminate between components of a mixture model. First, we generalize Eq.~\ref{eq:hepGen} to a mixture model of several components
\begin{equation}
p(x|\theta)=\sum_c w_c(\theta) p_c(x| \theta) \;.
\end{equation}
It is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems. 
\begin{eqnarray}\label{eq:decompose}
\frac{p(x|\theta_0)}{p(x|\theta_1)} & =& \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} 
\\
%&=& \sum_c \frac{ w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} \\
%&=& \sum_c \left[ \frac{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)}{ w_c(\theta_0) p_c(x| \theta_0)}  \right]^{-1} \\
&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(x| \theta_1)}{  p_c(x| \theta_0)}  \right]^{-1} \\ \label{eq:decomposedResult}
&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ p_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} 
\end{eqnarray}
The second line is a trivial, but useful decomposition into pair-wise classification between $p_{c'}(x|\theta_1)$ and $p_c(x|\theta_0)$.  The third line uses Theorem~1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$. In the situation where the only free parameters of the  model are the mixture coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before evaluating the  likelihood ratio). Equation~\ref{eq:decomposedResult} allows one to take advantage of both the parametrized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.

@@

Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\citek{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\citek{Allanach:2000kt}.  Both of these examples involve data sets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters. 
The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citek{Gao:2010qx,Bolognesi:2012mm}.

@@
